---
title: "Take-home Exercise 2: Application of Spatial and Spatio-temporal Analysis Methods to Discover the Distribution of Dengue Fever in Tainan City, Taiwan"
author:
  - name: Felicia Eng
date: "February 2024"
date-modified: "`r Sys.Date()`"
execute: 
  echo: true
  eval: true
  warning: false
editor: visual
---

## **The Task**

The specific tasks of this take-home exercise are as follows:

-   Using appropriate function of **sf** and **tidyverse**, preparing the following geospatial data layer:

    -   a study area layer in sf polygon features. It must be at village level and confined to the D01, D02, D04, D06, D07, D08, D32 and D39 counties of Tainan City, Taiwan.

    -   a dengue fever layer within the study area in sf point features. The dengue fever cases should be confined to epidemiology week 31-50, 2023.

    -   a derived dengue fever layer in [spacetime s3 class of sfdep](https://sfdep.josiahparry.com/articles/spacetime-s3). It should contain, among many other useful information, a data field showing number of dengue fever cases by village and by epidemiology week.

-   Using the extracted data, perform global spatial autocorrelation analysis by using [sfdep methods](https://is415-gaa-tskam.netlify.app/in-class_ex/in-class_ex05/in-class_ex05-glsa).

-   Using the extracted data, perform local spatial autocorrelation analysis by using [sfdep methods](https://r4gdsa.netlify.app/chap10.html).

-   Using the extracted data, perform emerging hotspot analysis by using [sfdep methods](https://is415-gaa-tskam.netlify.app/in-class_ex/in-class_ex05/in-class_ex05-ehsa).

-   Describe the spatial patterns revealed by the analysis above.

## **Importing Packages**

```{r}
pacman::p_load(sf, st, tidyverse, lubridate, spdep, tmap, knitr)
```

## **Importing Datasets to R Environment**

```{r}
tainan <- st_read(dsn = "data/geospatial", 
                 layer = "TAINAN_VILLAGE") 
```

```{r}
dengue <- read_csv("data/aspatial/Dengue_Daily.csv")
```






