---
title: "Hands on Exercise 3: 1st Order Spatial Point Patterns Analysis Methods"
---

# **1. Installing and Loading the R packages**

In this hands-on exercise, five R packages will be used, they are:

-   [**sf**](https://r-spatial.github.io/sf/), a relatively new R package specially designed to import, manage and process vector-based geospatial data in R.

-   [**spatstat**](https://spatstat.org/), which has a wide range of useful functions for point pattern analysis. In this hands-on exercise, it will be used to perform 1st- and 2nd-order spatial point patterns analysis and derive kernel density estimation (KDE) layer.

-   [**raster**](https://cran.r-project.org/web/packages/raster/) which reads, writes, manipulates, analyses and model of gridded spatial data (i.e. raster). In this hands-on exercise, it will be used to convert image output generate by spatstat into raster format.

-   [**maptools**](https://cran.r-project.org/web/packages/maptools/) which provides a set of tools for manipulating geographic data. In this hands-on exercise, we mainly use it to convert *Spatial* objects into *ppp* format of **spatstat**.

-   [**tmap**](https://cran.r-project.org/web/packages/tmap/) which provides functions for plotting cartographic quality static point patterns maps or interactive maps by using [leaflet](https://leafletjs.com/) API.

Use the code chunk below to install and launch the five R packages.

```{r}
install.packages("maptools", repos = "https://packagemanager.posit.co/cran/2023-10-13")
```

```{r}
pacman::p_load(maptools, sf, raster, spatstat, tmap)
```

> [Own notes:]{.underline}
>
> -   ppp is a point pattern object (of spatstat package)
>
>     Difference between Spatial Data Types (Vector and Raster):
>
> -   Raster data works with pixels
>
> -   Vector data consists of coordinates
>
> -   we can convert vector to raster, vice versa

# **2. Spatial Data Wrangling**

## **2.1 Importing the spatial data**

In this section, [*st_read()*](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package will be used to import these three geospatial data sets into R.

-   <div>

    Own notes:

    > -   **Spatial vs Geospatial**: geospatial is of or pertaining to a geographic location, especially data while spatial is of or pertaining to space
    >
    > -   **st_transform**: takes a geometry (childcare_sf) and a spatial reference system identifier (3414) as input parameters and transforms the geometry to be represented in the given spatial reference system

    </div>

```{r}
childcare_sf <- st_read("data/child-care-services-geojson.geojson") %>%
  st_transform(crs = 3414)
```

```{r}
sg_sf <- st_read(dsn = "data", layer="CostalOutline")
```

```{r}
mpsz_sf <- st_read(dsn = "data", 
                layer = "MP14_SUBZONE_WEB_PL")
```

1)  **DIY: Using the appropriate sf function you learned in Hands-on Exercise 2, retrieve the referencing system information of these geospatial data.**

```{r}
st_crs(childcare_sf)
```

```{r}
st_crs(mpsz_sf)
```

```{r}
st_crs(sg_sf)
```

2)  **DIY: Using the method you learned in Lesson 2, assign the correct crs to mpsz_sf and sg_sf simple feature data frames.**

```{r}
mpsz_sf <- st_transform(mpsz_sf, crs= 3414)
sg_sf <- st_transform(sg_sf, crs= 3414)
st_crs(mpsz_sf)
```

```{r}
st_crs(sg_sf)
```

3)  **DIY: Using the mapping methods you learned in Hands-on Exercise 3, prepare a map as shown below.**

```{r}
tm_shape(sg_sf) +
  tm_polygons() +
tm_shape(mpsz_sf) +
  tm_polygons() +
tm_shape(childcare_sf)+
  tm_dots()
```

Notice that all the geospatial layers are within the same map extend. This shows that their referencing system and coordinate values are referred to similar spatial context. This is very important in any geospatial analysis.

Alternatively, we can also prepare a pin map by using the code chunk below.

```{r}
tmap_mode('view')
tm_shape(childcare_sf)+
  tm_dots()
```

```{r}
tmap_mode('plot')
```

Notice that at the interactive mode, **tmap** is using [**leaflet for R**](https://rstudio.github.io/leaflet/) API. The advantage of this interactive pin map is it allows us to navigate and zoom around the map freely. We can also query the information of each simple feature (i.e. the point) by clicking of them. Last but not least, you can also change the background of the internet map layer. Currently, three internet map layers are provided. They are: ESRI.WorldGrayCanvas, OpenStreetMap, and ESRI.WorldTopoMap. The default is ESRI.WorldGrayCanvas.

> Reminder: Always remember to switch back to plot mode after the interactive map. This is because, each interactive mode will consume a connection. You should also avoid displaying ecessive numbers of interactive maps (i.e. not more than 10) in one RMarkdown document when publish on Netlify.

# 3. Data Wrangling

Although simple feature data frame is gaining popularity again sp’s Spatial\* classes, there are, however, many geospatial analysis packages require the input geospatial data in sp’s Spatial\* classes. In this section, you will learn how to convert simple feature data frame to sp’s Spatial\* class.

## 3.1 Converting sf data frames to sp’s Spatial\* class

The code below uses [*as_Spatial()*](https://r4gdsa.netlify.app/chap04.html) of **sf** package to convert the three geospatial data from simple feature data frame to sp’s Spatial\* class.

```{r}
childcare <- as_Spatial(childcare_sf)
mpsz <- as_Spatial(mpsz_sf)
sg <- as_Spatial(sg_sf)
```

DIY:Using appropriate function, display the information of these three Spatial\* classes as shown below.

```{r}
childcare
```

```{r}
mpsz
```

```{r}
sg
```

### 3.1.1 Seeing the difference between sp and sf (childcare)

```{r}
head(childcare_sf)
```

```{r}
head(childcare)
```

```{r}
plot(childcare_sf)
```

```{r}
plot(childcare)
```

```{r}
head(mpsz_sf)
```

```{r}
head(mpsz)
```

```{r}
plot(mpsz_sf)
```

```{r}
plot(mpsz)
```

```{r}
head(sg)
```

```{r}
plot(sg_sf)
```

```{r}
plot(sg)
```

## 3.2 Converting the Spatial\* class into generic sp format

spatstat requires the analytical data in ppp object form. There is no direct way to convert a Spatial\* classes into ppp object. We need to convert the Spatial classes\* into Spatial object first.

This code below will be converting Spatial\* classes into generic sp objects

```{r}
childcare_sp <- as(childcare, "SpatialPoints")
sg_sp <- as(sg, "SpatialPolygons")
```

To display the sp objects properties

```{r}
childcare_sp
```

```{r}
sg_sp
```

> Challenge: Do you know what are the differences between Spatial\* classes and generic sp object?

-   You can see that the previous SpatialPolygonsDataFrame class has now changed to the respective Spatial objects (points/polygons)

-   The points in a SpatialPoints object may be associated with a row of attributes to create a SpatialPointsDataFrame object

## 3.3 Converting the generic sp format into spatstat’s ppp format

we will use *as.ppp()* function of **spatstat** to convert the spatial data into **spatstat**’s ***ppp*** object format.

```{r}
childcare_ppp <- as(childcare_sp, "ppp")
childcare_ppp
```

Let's examine the difference by plotting chidlcare_ppp

```{r}
plot(childcare_ppp)
```

We now look at the summary statistics of the newly created ppp object.

```{r}
summary(childcare_ppp)
```

> **Key Point:** Pay attention to the warning message regarding duplicates. In spatial point patterns analysis, a notable concern is the existence of duplicates. The statistical methods applied to spatial point patterns are predominantly built on the assumption that processes are straightforward, meaning that points cannot overlap.

## 3.4 Handling duplicated points

In order to check dupication in a ppp object:

```{r}
any(duplicated(childcare_ppp))
```

multiplicity() function is used to count the number of co-indicence point.

```{r}
multiplicity(childcare_ppp)
```

If we want to know how many locations have more than one point event:

```{r}
sum(multiplicity(childcare_ppp) > 1)
```

Hence, it shows there are 128 duplicated point events.

To view the locations of these duplicate point events, we will plot childcare data:

```{r}
tmap_mode('view')
tm_shape(childcare) +
  tm_dots(alpha=0.4, 
          size=0.05)
```

```{r}
tmap_mode('plot')
```

Challenge: How to spot the duplicate points from the map shown above?

### 3.4.1 Jittering

```{r}
childcare_ppp_jit <- rjitter(childcare_ppp, 
                             retry=TRUE, 
                             nsim=1, 
                             drop=TRUE)
```

DIY: Using the method you learned in previous section, check if any dusplicated point in this geospatial data.

```{r}
any(duplicated(childcare_ppp_jit))
```

# 3.5 Creating owin object

When analysing spatial point patterns, it is a good practice to confine the analysis with a geographical area like Singapore boundary. In spatstat, an object called owin is specially designed to represent this polygonal region.

How to covert sg SpatialPolygon object into owin object of spatstat.

```{r}
sg_owin <- as(sg_sp, "owin")
```

The output object can be displayed by using plot() function

```{r}
plot(sg_owin)
```

summary() function of Base R

```{r}
summary(sg_owin)
```

## 3.6 Combining point events object and owin object

The last step of geospatial data wrangling, we will extract childcare events that are located within Singapore by using the code chunk below.

```{r}
childcareSG_ppp = childcare_ppp[sg_owin]
```

The output object combined both the point and polygon feature in one ppp object class as shown below.

```{r}
summary(childcareSG_ppp)
```

DIY: USING THE METHOD YOU LEARNED IN PREVIOUS EXERCISE, PLOT THE NEWLY DERIVED CHILDCARESG_PPP.

```{r}
plot(childcareSG_ppp)
```

# 4. First-order Spatial Point Patterns Analysis

In this section, you will learn how to perform first-order SPPA by using **spatstat** package. The hands-on exercise will focus on:

-   deriving **kernel density estimation (KDE)** layer for visualising and exploring the intensity of point processes,

-   performing **Confirmatory Spatial Point Patterns Analysis** by using **Nearest Neighbour** statistics.

## 4.1 Kernel Density Estimation

In this section, you will learn how to compute the kernel density estimation (KDE) of childcare services in Singapore.

#### 4.1.1 Computing kernel density estimation using automatic bandwidth selection method

The code chunk below computes a kernel density by using the following configurations of [*density()*](https://rdrr.io/cran/spatstat/man/density.ppp.html) of **spatstat**:

-   [*bw.diggle()*](https://rdrr.io/cran/spatstat/man/bw.diggle.html) automatic bandwidth selection method. Other recommended methods are [*bw.CvL()*](https://rdrr.io/cran/spatstat/man/bw.CvL.html), [*bw.scott()*](https://rdrr.io/cran/spatstat/man/bw.scott.html) or [*bw.ppl()*](https://rdrr.io/cran/spatstat/man/bw.ppl.html).

-   The smoothing kernel used is *gaussian*, which is the default. Other smoothing methods are: “epanechnikov”, “quartic” or “disc”.

-   The intensity estimate is corrected for edge effect bias by using method described by Jones (1993) and Diggle (2010, equation 18.9). The default is *FALSE*.

```{r}
kde_childcareSG_bw <- density(childcareSG_ppp,
                              sigma=bw.diggle,
                              edge=TRUE,
                            kernel="gaussian") 
```

The plot() function of Base R is then used to display the kernel density derived.

```{r}
plot(kde_childcareSG_bw)
```

The density values of the output range from 0 to 0.000035 which is way too small to comprehend. This is because the default unit of measurement of svy21 is in meter. As a result, the density values computed is in “number of points per square meter”.

Before we move on to next section, it is good to know that you can retrieve the bandwidth used to compute the kde layer:

```{r}
bw <- bw.diggle(childcareSG_ppp)
bw
```

### 4.1.2 Rescalling KDE values

rescale() is used to covert the unit of measurement from meter to kilometer:

```{r}
childcareSG_ppp.km <- rescale(childcareSG_ppp, 1000, "km")
```

Now, we can re-run density() using the resale data set and plot the output kde map.

```{r}
kde_childcareSG.bw <- density(childcareSG_ppp.km, sigma=bw.diggle, edge=TRUE, kernel="gaussian")
plot(kde_childcareSG.bw)
```

Now we notice the output image looks identical to the earlier version, the only changes in the data values (refer to the legend).

## 4.2 Working with different automatic badwidth methods

Beside bw.diggle(), there are three other spatstat functions can be used to determine the bandwidth, they are: bw.CvL(), bw.scott(), and bw.ppl().

Let us take a look at the bandwidth return by these automatic bandwidth calculation methods by using:

```{r}
 bw.CvL(childcareSG_ppp.km)
```

```{r}
bw.scott(childcareSG_ppp.km)
```

```{r}
bw.ppl(childcareSG_ppp.km)
```

```{r}
bw.diggle(childcareSG_ppp.km)
```

Baddeley et. (2016) suggested the use of the bw.ppl() algorithm because in ther experience it tends to produce the more appropriate values when the pattern consists predominantly of tight clusters. But they also insist that if the purpose of once study is to detect a single tight cluster in the midst of random noise then the bw.diggle() method seems to work best.

This will be used to compare the output of using bw.diggle and bw.ppl methods.

```{r}
kde_childcareSG.ppl <- density(childcareSG_ppp.km, 
                               sigma=bw.ppl, 
                               edge=TRUE,
                               kernel="gaussian")
par(mfrow=c(1,2))
plot(kde_childcareSG.bw, main = "bw.diggle")
plot(kde_childcareSG.ppl, main = "bw.ppl")
```

## 4.3 Working with different kernel methods

By default, the kernel method used in density.ppp() is gaussian. But there are three other options, namely: Epanechnikov, Quartic and Dics. Let’s take a look at what they’re like:

```{r}
par(mfrow=c(2,2))
plot(density(childcareSG_ppp.km, 
             sigma=bw.ppl, 
             edge=TRUE, 
             kernel="gaussian"), 
     main="Gaussian")
plot(density(childcareSG_ppp.km, 
             sigma=bw.ppl, 
             edge=TRUE, 
             kernel="epanechnikov"), 
     main="Epanechnikov")
plot(density(childcareSG_ppp.km, 
             sigma=bw.ppl, 
             edge=TRUE, 
             kernel="quartic"), 
     main="Quartic")
plot(density(childcareSG_ppp.km, 
             sigma=bw.ppl, 
             edge=TRUE, 
             kernel="disc"), 
     main="Disc")
```

## 4.4 Fixed and Adaptive KDE

### 4.4.1 Computing KDE by using fixed bandwidth

Next, we will compute a KDE layer by defining a bandwidth of 600 meter. Notice that in the code chunk below, the sigma value used is 0.6. This is because the unit of measurement of ***childcareSG_ppp.km*** object is in kilometer, hence the 600m is 0.6km.

```{r}
kde_childcareSG_600 <- density(childcareSG_ppp.km, sigma=0.6, edge=TRUE, kernel="gaussian")
plot(kde_childcareSG_600)
```

### 4.4.2 Computing KDE by using adaptive bandwidth

Fixed bandwidth method is very sensitive to highly skew distribution of spatial point patterns over geographical units for example urban versus rural. One way to overcome this problem is by using adaptive bandwidth instead.

In this section, we will learn how to derive adaptive kernel density estimation by using [*density.adaptive()*](https://rdrr.io/cran/spatstat/man/adaptive.density.html) of **spatstat**.

```{r}
kde_childcareSG_adaptive <- adaptive.density(childcareSG_ppp.km, method="kernel")
plot(kde_childcareSG_adaptive)
```

We can compare the fixed and adaptive kernel density estimation outputs by using:

```{r}
par(mfrow=c(1,2))
plot(kde_childcareSG.bw, main = "Fixed bandwidth")
plot(kde_childcareSG_adaptive, main = "Adaptive bandwidth")
```

### 4.4.3 Converting KDE output into grid object

The result is the same, we just convert it so that it is suitable for mapping purposes

```{r}
gridded_kde_childcareSG_bw <- as.SpatialGridDataFrame.im(kde_childcareSG.bw)
spplot(gridded_kde_childcareSG_bw)
```

#### 4.4.3.1 Converting gridded output into raster

Next, we will convert the gridded kernal density objects into RasterLayer object by using raster() of raster package.

```{r}
kde_childcareSG_bw_raster <- raster(gridded_kde_childcareSG_bw)
```

Let us take a look at the properties of kde_childcareSG_bw_raster RasterLayer.

```{r}
kde_childcareSG_bw_raster
```

#### 4.4.3.2 Assigning projection systems

Below we will be used to include the CRS information on kde_childcareSG_bw_raster RasterLayer.

```{r}
projection(kde_childcareSG_bw_raster) <- CRS("+init=EPSG:3414")
kde_childcareSG_bw_raster
```

Notice: crs property is completed

### 4.4.4 Visualising the output in tmap

Finally, we will display the raster in cartographic quality map using tmap package.

```{r}
tm_shape(kde_childcareSG_bw_raster) + 
  tm_raster("v") +
  tm_layout(legend.position = c("right", "bottom"), frame = FALSE)
```

Notice: The raster values are encoded explicitly onto the raster pixel using the values in “v”” field.

### 4.4.5 Comparing Spatial Point Patterns using KDE

In this section, we will learn how to compare KDE of childcare at Punggol, Tampines, Chua Chu Kang and Jurong West planning areas.

#### 4.4.5.1 Extracting study area

To extract the target planning areas:

```{r}
pg = mpsz[mpsz@data$PLN_AREA_N == "PUNGGOL",]
tm = mpsz[mpsz@data$PLN_AREA_N == "TAMPINES",]
ck = mpsz[mpsz@data$PLN_AREA_N == "CHOA CHU KANG",]
jw = mpsz[mpsz@data$PLN_AREA_N == "JURONG WEST",]
```

Plotting target planning areas

```{r}
par(mfrow=c(2,2))
plot(pg, main = "Ponggol")
plot(tm, main = "Tampines")
plot(ck, main = "Choa Chu Kang")
plot(jw, main = "Jurong West")
```

#### 4.4.5.2 Converting the spatial point data frame into generic sp format

we will convert these SpatialPolygonsDataFrame layers into generic spatialpolygons layers.

```{r}
pg_sp = as(pg, "SpatialPolygons")
tm_sp = as(tm, "SpatialPolygons")
ck_sp = as(ck, "SpatialPolygons")
jw_sp = as(jw, "SpatialPolygons")
```

#### 4.4.5.3 Creating owin object

Now, we will convert these SpatialPolygons objects into owin objects that is required by spatstat.

```{r}
pg_owin = as(pg_sp, "owin")
tm_owin = as(tm_sp, "owin")
ck_owin = as(ck_sp, "owin")
jw_owin = as(jw_sp, "owin")
```

#### 4.4.5.4 Combining childcare points and the study area

To extract childcare that is within the specific region to do our analysis later on.

```{r}
childcare_pg_ppp = childcare_ppp_jit[pg_owin]
childcare_tm_ppp = childcare_ppp_jit[tm_owin]
childcare_ck_ppp = childcare_ppp_jit[ck_owin]
childcare_jw_ppp = childcare_ppp_jit[jw_owin]
```

Next, rescale() function is used to trasnform the unit of measurement from metre to kilometre.

```{r}
childcare_pg_ppp.km = rescale(childcare_pg_ppp, 1000, "km")
childcare_tm_ppp.km = rescale(childcare_tm_ppp, 1000, "km")
childcare_ck_ppp.km = rescale(childcare_ck_ppp, 1000, "km")
childcare_jw_ppp.km = rescale(childcare_jw_ppp, 1000, "km")
```

To plot these four study areas and the locations of the childcare centres.

```{r}
par(mfrow=c(2,2))
plot(childcare_pg_ppp.km, main="Punggol")
plot(childcare_tm_ppp.km, main="Tampines")
plot(childcare_ck_ppp.km, main="Choa Chu Kang")
plot(childcare_jw_ppp.km, main="Jurong West")
```

#### 4.4.5.5 Computing KDE

To compute the KDE of these four planning area. bw.diggle method is used to derive the bandwidth of each

```{r}
par(mfrow=c(2,2))
plot(density(childcare_pg_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Punggol")
plot(density(childcare_tm_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Tempines")
plot(density(childcare_ck_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Choa Chu Kang")
plot(density(childcare_jw_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="JUrong West")
```

#### 4.4.5.6 Computing fixed bandwidth KDE

For comparison purposes, we will use 250m as the bandwidth.

```{r}
par(mfrow=c(2,2))
plot(density(childcare_ck_ppp.km, 
             sigma=0.25, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Chou Chu Kang")
plot(density(childcare_jw_ppp.km, 
             sigma=0.25, 
             edge=TRUE, 
             kernel="gaussian"),
     main="JUrong West")
plot(density(childcare_pg_ppp.km, 
             sigma=0.25, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Punggol")
plot(density(childcare_tm_ppp.km, 
             sigma=0.25, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Tampines")
```

## 4.5 Nearest Neighbour Analysis

In this section, we will perform the Clark-Evans test of aggregation for a spatial point pattern by using [*clarkevans.test()*](https://www.rdocumentation.org/packages/spatstat/versions/1.63-3/topics/clarkevans.test) of **statspat**.

The test hypotheses are:

Ho = The distribution of childcare services are randomly distributed.

H1= The distribution of childcare services are not randomly distributed.

The 95% confident interval will be used.

### 4.5.1 Testing spatial point patterns using Clark and Evans Test

```{r}
clarkevans.test(childcareSG_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                nsim=99)
```

### 4.5.2 Clark and Evans Test: Choa Chu Kang planning area

In the code chunk below, clarkevans.test() of spatstat is used to performs Clark-Evans test of aggregation for childcare centre in Choa Chu Kang planning area.

```{r}
clarkevans.test(childcare_ck_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)
```

### 4.5.3 Clark and Evans Test: Tampines planning area

The similar test is used to analyse the spatial point patterns of childcare centre in Tampines planning area.

```{r}
clarkevans.test(childcare_tm_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)
```


